---
title: "Plot_Fst"
author: "Melanie LaCava"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Calculate pairwise Fst values using ANGSD

We're using ANGSD to calculate Fst among our sites, following this guide: http://www.popgen.dk/angsd/index.php/Fst.

You must calculate the unfolded site frequency spectrum for each of your sites before you can calculate pairwise Fst values - before running this script get_fst.sh, run get_div_all.sh or get_div_singlepop.sh to produce the site frequency spectrum files you need as input here.

```{bash, eval=FALSE}
#!/bin/bash
#SBATCH -t 01:00:00
#SBATCH --job-name=fst
#SBATCH -p high
#SBATCH --output=fst-%A-%a.out # File to which STDOUT will be written
#SBATCH --error=fst-%A-%a.err # File to which STDERR will be written
#SBATCH --mail-type=END,FAIL
#SBATCH --mail-user=mlacava@ucdavis.edu

#RUN get_div_all.sh or get_div_singlepop.sh before running this script

#MODIFY code below in 2 ways:
# - provide path to your existing .saf.idx files from diversity stats
# - insert your username to submit each job under your account

#SUBMIT with: sbatch get_fst.sh [pop list]
# - pop list is the list of site codes you want to calc Fst among (this script will loop throgh all possible pairs of these sites and create and submit an SBATCH job to run Fst for them) (e.g., all.pop.list)

module load angsd

infile=$1 #your pop list (e.g., all.pop.list)
n=$(wc -l $infile | awk '{print $1}')

x=1 
while [ $x -le $n ]
do
    y=$(( $x + 1 ))
    while [ $y -le $n ]
    do
    
    pop1=$( (sed -n ${x}p $infile) )
    pop2=$( (sed -n ${y}p $infile) )

        echo "#!/bin/bash" > ${pop1}.${pop2}.sh
        echo "" >> ${pop1}.${pop2}.sh
        #change relative or absolute path to your existing .saf.idx files below
        echo "realSFS ../diversity/${pop1}.saf.idx ../diversity/${pop2}.saf.idx > ${pop1}.${pop2}.2dsfs" >> ${pop1}.${pop2}.sh
        echo "" >> ${pop1}.${pop2}.sh
        echo "realSFS fst index ../diversity/${pop1}.saf.idx ../diversity/${pop2}.saf.idx -sfs ${pop1}.${pop2}.2dsfs -fstout ${pop1}.${pop2}" >> ${pop1}.${pop2}.sh
        echo "" >> ${pop1}.${pop2}.sh
        echo "realSFS fst stats ${pop1}.${pop2}.fst.idx 2> ${pop1}.${pop2}_global.fst" >> ${pop1}.${pop2}.sh

        #insert your FARM username below to submit sbatch job, increase time requested if your jobs don't finish
        sbatch -J mlacava -p med --mem=32G -t 06:00:00 -c 1 ${pop1}.${pop2}.sh

    y=$(( $y + 1 ))
    
    done

x=$(( $x + 1 ))

done
```



## Compile Fst values from output files

```{bash, eval=FALSE}

#Just run in command line on FARM in the folder with your Fst output files to grab the pop names and unweighted and weighted Fst values for each pop pair

echo "pop1,pop2,UnweightFst,WeightFst" > compiledFsts.csv #create csv with desired headers
for i in *global.fst; do
pop1=$(sed -n 1p $i | awk -F':' '{print $2}'  | awk -F'.' '{print $1}')
pop2=$(sed -n 1p $i | awk -F':' '{print $2}'  | awk -F'.' '{print $2}')
uFst=$(sed -n 3p $i | awk -F':' '{print $3}' | awk -F' ' '{print $1}')
wFst=$(sed -n 3p $i | awk -F':' '{print $4}')
echo "$pop1,$pop2,$uFst,$wFst" >> compiledFsts.csv; done

```


## Fst tile plot

Plot pairwise Fst values for your sites in a ggplot tile plot 

```{r, eval=FALSE}
#import compiled global Fsts
fst <- read.csv("https://raw.githubusercontent.com/meflacava/RelictDace/main/results/compiledFsts.csv")
head(fst)
names(fst) <- c("pop1","pop2","UnweightFst","Fst") #change weighted Fst name for legend plotting

#select subset of site pairs to plot (e.g., only western valley sites)
west <- read.table("https://raw.githubusercontent.com/meflacava/RelictDace/main/bamlists/west.pop.list")
west <- west$V1
fst <- fst[fst$pop1 %in% west,]

#tile plot
ggplot(fst, aes(pop2, pop1)) + 
  geom_tile(aes(fill = Fst), color="white") + 
  theme_minimal() + 
  scale_fill_gradient(low="#98afd4",high="#04255c") + #blue gradient
  #scale_fill_viridis_c(direction=-1,option = "cividis") + #Finger 2022 manuscript colors
  theme_bw() + 
  theme(axis.text.x=element_text(angle=45,vjust=1,size=10,hjust=1), panel.border=element_blank(),
        panel.grid.major=element_blank(), axis.title.y = element_blank(), axis.title.x = element_blank()) + scale_y_discrete(position = "left") + 
  geom_text(aes(label = round(Fst, 2)), size=3.5, colour = "white") + 
  #ggtitle("D") + 
  theme(plot.title = element_text(face="bold", size=16))
```